### 2021-04-19 paper/FAQ_STA561_Project.Rmd
GetPossiblePatterns function

Most record linkage techniques are derived from the seminal 1969 paper by Fellegi and Sunter: "A Theory for Record Linkage". The defining characteristic of their model was to transform the sets of records, which constitute text data that is difficult to model, into sets of comparison vectors governed by parameters that can be more easily estimated. Concretely, if files $A$ and $B$ have $n_A$ and $n_B$ records respectively, and if the files share $F$ fields in common upon which to base the linkage, the Fellegi and Sunter approach generates a $n_A n_B \times F$ matrix $\Gamma$, which contains similarity scores between each pair of records across datasets. We say $\gamma_{ij}$ is the comparison vector for record $i \in A$ and record $j \in B$, with $\gamma_{ij}^f$ providing their similarity score on the $f^{th}$ field. For ease of modeling and computation, we restrict these similarity scores to be discrete, ordinal variables, and the construction of these is left to the modeller. With notation $\gamma_{ij}^f \in \{0, \ldots, L_f - 1\}$ for the number of possible agreement levels for field $f$, we use binary 0-1 variables to indicate exact matching, and 0-1-2 variables to provide an option for partial matching. For text data, we calculate similarity based on Levenstein distance or some other text similarity score, and bin these scores to integers for use in the model.

@@

Broadly speaking, we increase our computational efficiency by recognizing that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. Let $H$ be the set of unique agreement patterns in the data, let $P$ denote the total number of unique agreement patterns. Note that $P$ is bounded above by $\prod_{f=1}^F L_f$, and that this bound does not scale with $n_A$ or $n_B$. We index these agreement patterns by $p \in \{1, \ldots, P\}$, and say $(i,j) \in h_p$ when the $(i,j)$ pair exhibits the $p^{th}$ agreement pattern. Wherever possible, we conduct calculations over these $P$ agreement patterns rather than the $n_A \times n_B$ record pairs.

By making linkage decisions for each $j\in B$ independently, we do not need to concern ourselves with the precise record label for $j's$ potential pairing within the Gibbs sampler, only the agreement pattern of the $\gamma$ vector. While the original Fellegi Sunter framework has computational complexity $O(n_A \times n_B)$, these modifications completely remove dependence on $n_A$ throughout the Gibbs sampler, reducing the complexity to $O(n_B)$

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by running `BRL` on simulated pairs of files with differing levels of errors and overlap. Comparing our method against `BRL` on these same simulated datasets, we find that our method only has weakend performance in the extreme scenario of very high errors and very high overlap across files.

$$\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}$$
$$\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}$$
$$\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)$$

```{r fig.align='center'}
knitr::include_graphics("../notes/figures/sadinle_sim_plot.png")
```

@@

We simulate $\Gamma$ matrices of comparison vectors from multinomial distributions meant to emulate the behavior of similarity scores across first name, last name, and day, month, and and year of birth. Distributions provided below. We simulate these data for different values of $n_A and n_B$, and compare the runtime of `parlr` against `BRL`. We note here that `BRL` is coded in C, which makes for unfair comparison against `parlr`, currently only built in R. We see that at low data size, `BRL` outperforms, but that `parlr` is significantly faster at handling larger data. In particular, runtime for `BRL` seems to grow quadratically while runtime for `parlr` seems to grow linearly. Additionally, although `parlr` is amenable to parallezation, this simulation was run on a single core. Running `parlr` in C or C++ with paralellization for the hashing step and sampling the matching status of the record pairs should lead to even more drastic results.

```{r fig.align='center'}
knitr::include_graphics("../notes/figures/sadinle_speed_plot.png")
```

The above discussion suggests that for fixed $n_B$, computation time should remain mostly constant with growing $n_A$. Shockingly, this seems to be true. In the plot below, fixing $n_B = 500$, we see linear growth for the runtime under `BRL` as $n_A$ increases, with much more static runtime under `parlr`. The slight increases in runtime that we do see are due primarily to the hashing step, which again can be run in parallel for large data.

```{r fig.align='center'}
knitr::include_graphics("../notes/figures/speed_plot_fixed_nB.png")
```

@@

At the moment, `parlr` (and its `BRL` competitor) are only able to handle bipartite matching across two files, meaning there is duplication across, but not within, files. Though it seems similar to linkage across files, deduplication within one file is suprisingly much more difficult to model probabalistically. In particular, `parlr` assumes  that the linkage decision for record $j \in B$ is independent from all other $j' \in B$, allowing us to propose a prior $\lambda \sim \text{Beta}(\alpha, \beta)$ for the probability of $j$ having some match in $A$, and update that prior through standard beta-binomial posterior updates. With only one file however, if the sampler decides that record $1$ matches with record $5$, there is no way to determine the linkage of record $5$ independently; we do not even have $n_B$ independent linkage decisions to work with! Constructing fast, scalable methods for deduplication is one avenue of future work.

>>>>

Most record linkage techniques are derived from the seminal 1969 paper by Fellegi and Sunter: "A Theory for Record Linkage". The defining characteristic of their model was to transform the sets of records, which constitute text data that is difficult to model, into sets of comparison vectors governed by parameters that can be more easily estimated. Concretely, if files $A$ and $B$ have $n_A$ and $n_B$ records respectively, and if the files share $F$ fields in common upon which to base the linkage, the Fellegi and Sunter approach generates a $n_A n_B \times F$ matrix $\Gamma$, which contains similarity scores between each pair of records across datasets. We say $\gamma_{ij}$ is the comparison vector for record $i \in A$ and record $j \in B$, with $\gamma_{ij}^f$ providing their similarity score on the $f^{th}$ field. For ease of modeling and computation, we restrict these similarity scores to be discrete, ordinal variables, and the construction of these is left to the modeller. It is common to use use binary 0-1 variables to indicate exact matching, and 0-1-2 variables to provide an option for partial matching. For text data, we calculate similarity based on Levenstein distance or some other text similarity score, and bin these scores to integers for use in the model.

@@

Broadly speaking, we increase our computational efficiency by recognizing that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. Let $P$ denote the total number of unique agreement patterns; note that $P$ is bounded above by $\prod_{f=1}^F L_f$, and that this bound does not scale with $n_A$ or $n_B$. We index these agreement patterns by $p \in \{1, \ldots, P\}$, and say $(i,j) \in h_p$ when the $(i,j)$ pair exhibits the $p^{th}$ agreement pattern. Wherever possible, we conduct calculations over these $P$ agreement patterns rather than the $n_A \times n_B$ record pairs. While the original Fellegi Sunter framework has computational complexity $O(n_A \times n_B)$, these modifications completely remove dependence on $n_A$ throughout the Gibbs sampler, reducing the complexity to $O(n_B)$

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by running `BRL` on simulated pairs of files with differing levels of errors and overlap. Comparing our method against `BRL` on these same simulated datasets, we find that our method only has weakend performance in the extreme scenario of very high errors and very high overlap across files. See plots in technical appendix.

@@

We simulate $\Gamma$ matrices of comparison vectors from multinomial distributions meant to emulate the behavior of similarity scores across first name, last name, and day, month, and and year of birth. We simulate these data for different values of $n_A$ and $n_B$, and compare the runtime of `parlr` against `BRL`. We see that at low data size, `BRL` outperforms, but that `parlr` is significantly faster at handling larger data. See plots in technical appendix.

### How fast can it be?

We note here that `BRL` is coded in C, which makes for unfair comparison against `parlr`, currently only built in R. Additionally, although `parlr` is amenable to parallezation, our simulations thus far have only been run on a single core. Running `parlr` in C++ with paralellization for the hashing step and sampling the matching status of the record pairs should lead to even more drastic results.

@@

At the moment, `parlr` (and its `BRL` competitor) is only able to handle bipartite matching across two files, meaning there is duplication across, but not within, files. Though it seems similar to linkage across files, deduplication within one file is suprisingly much more difficult to model probabalistically. In particular, `parlr` assumes  that the linkage decision for record $j \in B$ is independent from all other $j' \in B$, allowing us to propose a prior $\lambda \sim \text{Beta}(\alpha, \beta)$ for the probability of $j$ having some match in $A$, and update that prior through standard beta-binomial posterior updates. However, with only one file of $n$ records, we do not even have $n$ independent linkage decisions to work with! For example, if the sampler decides that record $1$ matches with record $5$, there is no way to determine the linkage of record $5$ independently.  Constructing fast, scalable methods for deduplication is one avenue of future work.

### 2021-04-19 paper/FAQ_STA561_Project.tex
GetPossiblePatterns function

Most record linkage techniques are derived from the seminal 1969 paper
by Fellegi and Sunter: ``A Theory for Record Linkage''. The defining
characteristic of their model was to transform the sets of records,
which constitute text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates a
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f\) providing their similarity score on the
\(f^{th}\) field. For ease of modeling and computation, we restrict
these similarity scores to be discrete, ordinal variables, and the
construction of these is left to the modeller. With notation
\(\gamma_{ij}^f \in \{0, \ldots, L_f - 1\}\) for the number of possible
agreement levels for field \(f\), we use binary 0-1 variables to
indicate exact matching, and 0-1-2 variables to provide an option for
partial matching. For text data, we calculate similarity based on
Levenstein distance or some other text similarity score, and bin these
scores to integers for use in the model.

@@

Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let \(H\)
be the set of unique agreement patterns in the data, let \(P\) denote
the total number of unique agreement patterns. Note that \(P\) is
bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not
scale with \(n_A\) or \(n_B\). We index these agreement patterns by
\(p \in \{1, \ldots, P\}\), and say \((i,j) \in h_p\) when the \((i,j)\)
pair exhibits the \(p^{th}\) agreement pattern. Wherever possible, we
conduct calculations over these \(P\) agreement patterns rather than the
\(n_A \times n_B\) record pairs.

By making linkage decisions for each \(j\in B\) independently, we do not
need to concern ourselves with the precise record label for \(j's\)
potential pairing within the Gibbs sampler, only the agreement pattern
of the \(\gamma\) vector. While the original Fellegi Sunter framework
has computational complexity \(O(n_A \times n_B)\), these modifications
completely remove dependence on \(n_A\) throughout the Gibbs sampler,
reducing the complexity to \(O(n_B)\)

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by
running \texttt{BRL} on simulated pairs of files with differing levels
of errors and overlap. Comparing our method against \texttt{BRL} on
these same simulated datasets, we find that our method only has weakend
performance in the extreme scenario of very high errors and very high
overlap across files.
\[\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}\]
\[\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}\]
\[\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)\]

\begin{center}\includegraphics[width=29.17in]{../notes/figures/sadinle_sim_plot} \end{center}

@@

We simulate \(\Gamma\) matrices of comparison vectors from multinomial
distributions meant to emulate the behavior of similarity scores across
first name, last name, and day, month, and and year of birth.
Distributions provided below. We simulate these data for different
values of \(n_A and n_B\), and compare the runtime of \texttt{parlr}
against \texttt{BRL}. We note here that \texttt{BRL} is coded in C,
which makes for unfair comparison against \texttt{parlr}, currently only
built in R. We see that at low data size, \texttt{BRL} outperforms, but
that \texttt{parlr} is significantly faster at handling larger data. In
particular, runtime for \texttt{BRL} seems to grow quadratically while
runtime for \texttt{parlr} seems to grow linearly. Additionally,
although \texttt{parlr} is amenable to parallezation, this simulation
was run on a single core. Running \texttt{parlr} in C or C++ with
paralellization for the hashing step and sampling the matching status of
the record pairs should lead to even more drastic results.

\begin{center}\includegraphics[width=23.6in]{../notes/figures/sadinle_speed_plot} \end{center}

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Shockingly, this
seems to be true. In the plot below, fixing \(n_B = 500\), we see linear
growth for the runtime under \texttt{BRL} as \(n_A\) increases, with
much more static runtime under \texttt{parlr}. The slight increases in
runtime that we do see are due primarily to the hashing step, which
again can be run in parallel for large data.

\begin{center}\includegraphics[width=29.17in]{../notes/figures/speed_plot_fixed_nB} \end{center}

@@

At the moment, \texttt{parlr} (and its \texttt{BRL} competitor) are only
able to handle bipartite matching across two files, meaning there is
duplication across, but not within, files. Though it seems similar to
linkage across files, deduplication within one file is suprisingly much
more difficult to model probabalistically. In particular, \texttt{parlr}
assumes that the linkage decision for record \(j \in B\) is independent
from all other \(j' \in B\), allowing us to propose a prior
\(\lambda \sim \text{Beta}(\alpha, \beta)\) for the probability of \(j\)
having some match in \(A\), and update that prior through standard
beta-binomial posterior updates. With only one file however, if the
sampler decides that record \(1\) matches with record \(5\), there is no
way to determine the linkage of record \(5\) independently; we do not
even have \(n_B\) independent linkage decisions to work with!
Constructing fast, scalable methods for deduplication is one avenue of
future work.

>>>>

Most record linkage techniques are derived from the seminal 1969 paper
by Fellegi and Sunter: ``A Theory for Record Linkage''. The defining
characteristic of their model was to transform the sets of records,
which constitute text data that is difficult to model, into sets of
comparison vectors governed by parameters that can be more easily
estimated. Concretely, if files \(A\) and \(B\) have \(n_A\) and \(n_B\)
records respectively, and if the files share \(F\) fields in common upon
which to base the linkage, the Fellegi and Sunter approach generates a
\(n_A n_B \times F\) matrix \(\Gamma\), which contains similarity scores
between each pair of records across datasets. We say \(\gamma_{ij}\) is
the comparison vector for record \(i \in A\) and record \(j \in B\),
with \(\gamma_{ij}^f\) providing their similarity score on the
\(f^{th}\) field. For ease of modeling and computation, we restrict
these similarity scores to be discrete, ordinal variables, and the
construction of these is left to the modeller. It is common to use use
binary 0-1 variables to indicate exact matching, and 0-1-2 variables to
provide an option for partial matching. For text data, we calculate
similarity based on Levenstein distance or some other text similarity
score, and bin these scores to integers for use in the model.

@@

Broadly speaking, we increase our computational efficiency by
recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. Let \(P\)
denote the total number of unique agreement patterns; note that \(P\) is
bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not
scale with \(n_A\) or \(n_B\). We index these agreement patterns by
\(p \in \{1, \ldots, P\}\), and say \((i,j) \in h_p\) when the \((i,j)\)
pair exhibits the \(p^{th}\) agreement pattern. Wherever possible, we
conduct calculations over these \(P\) agreement patterns rather than the
\(n_A \times n_B\) record pairs. While the original Fellegi Sunter
framework has computational complexity \(O(n_A \times n_B)\), these
modifications completely remove dependence on \(n_A\) throughout the
Gibbs sampler, reducing the complexity to \(O(n_B)\)

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by
running \texttt{BRL} on simulated pairs of files with differing levels
of errors and overlap. Comparing our method against \texttt{BRL} on
these same simulated datasets, we find that our method only has weakend
performance in the extreme scenario of very high errors and very high
overlap across files. See plots in technical appendix.

@@

We simulate \(\Gamma\) matrices of comparison vectors from multinomial
distributions meant to emulate the behavior of similarity scores across
first name, last name, and day, month, and and year of birth. We
simulate these data for different values of \(n_A\) and \(n_B\), and
compare the runtime of \texttt{parlr} against \texttt{BRL}. We see that
at low data size, \texttt{BRL} outperforms, but that \texttt{parlr} is
significantly faster at handling larger data. See plots in technical
appendix.

\hypertarget{how-fast-can-it-be}{%
\subsubsection{How fast can it be?}\label{how-fast-can-it-be}}

We note here that \texttt{BRL} is coded in C, which makes for unfair
comparison against \texttt{parlr}, currently only built in R.
Additionally, although \texttt{parlr} is amenable to parallezation, our
simulations thus far have only been run on a single core. Running
\texttt{parlr} in C++ with paralellization for the hashing step and
sampling the matching status of the record pairs should lead to even
more drastic results.

@@

At the moment, \texttt{parlr} (and its \texttt{BRL} competitor) is only
able to handle bipartite matching across two files, meaning there is
duplication across, but not within, files. Though it seems similar to
linkage across files, deduplication within one file is suprisingly much
more difficult to model probabalistically. In particular, \texttt{parlr}
assumes that the linkage decision for record \(j \in B\) is independent
from all other \(j' \in B\), allowing us to propose a prior
\(\lambda \sim \text{Beta}(\alpha, \beta)\) for the probability of \(j\)
having some match in \(A\), and update that prior through standard
beta-binomial posterior updates. However, with only one file of \(n\)
records, we do not even have \(n\) independent linkage decisions to work
with! For example, if the sampler decides that record \(1\) matches with
record \(5\), there is no way to determine the linkage of record \(5\)
independently. Constructing fast, scalable methods for deduplication is
one avenue of future work.
